Challenges to the Two-Infall Scenario by Large Stellar Age Catalogs

### 2025-12-03 src/tex/ms.tex
Updates from Overleaf

\begin{figure}
    \centering
    \includegraphics[width=\onecolumn]{\figpath{gas_abundance_evolution.pdf}}
    \caption{The ISM abundance evolution at $R_{\rm gal}=8\kpc$ of three multi-zone models at different yield scales (see Table \ref{tab:yields}), each of which feature a major dilution event 9 Gyr ago.
    % The dark gray (light gray) shaded region plots the 68\textsuperscript{th} (95\textsuperscript{th}) percentile of APOGEE abundances in the Solar neighborhood ($7\leq R_{\rm gal}<9\kpc$, $|z|<0.5\kpc$) with NN ages from \citet{leung_variational_2023}. The black line (black points) plots the median (mode) of the abundance data in 1 Gyr-wide age bins, and the gray error bars along the bottom of each panel indicate the median age and abundance errors as a function of age. 
    \added{The 2-D histograms show the APOGEE stellar age--abundance distributions in the Solar neighborhood ($7\leq R_{\rm gal}<9\kpc$, $|z|<0.5\kpc$), adopting the \citet{leung_variational_2023} ages. The black points plot the mode of the abundance data in 1 Gyr-wide age bins, and the gray error bars along the bottom of each panel indicate the median age and abundance errors as a function of age.}
    The left-hand marginal panels show the predicted (blue, pink, and green) and observed (gray) stellar abundance distributions \added{in the same region}, which are boxcar-smoothed with a width of 0.05 dex for visual clarity.}
    \label{fig:yield-outflow}
\end{figure}

Figure \ref{fig:yield-outflow} shows the mode of the APOGEE [O/H], [Fe/H], and [O/Fe] distributions as a function of age at the Solar annulus, using the NN ages from \citet{leung_variational_2023}. As shown by \citet{johnson_milky_2025}, the peak of the MDF is less susceptible to modification by radial migration, making it a more reliable proxy for ISM chemistry at a given lookback time than the mean or median.
We over-plot the ISM abundance evolution in our $y/Z_\odot=1,2,$ and $3$ models with the fiducial parameters at $R_{\rm gal}=8\kpc$ for comparison. The models are in broad agreement with each other and with the data at lookback times of $\lesssim5\Gyr$. At ages of $\sim5-8\Gyr$, the models underpredict the metallicities of the observed stars by up to half an order of magnitude, a natural consequence of the dilution associated with the second infall event. This discrepancy is modestly reduced in the higher $y/Z_\odot$ models, which re-enrich more rapidly. However, these models exacerbate tensions with the age--[O/Fe] relation because the O abundance increases more before enrichment from second-infall SNe Ia becomes important, making the $\alpha$-enhancement from the ensuing starburst more pronounced. Overall, none of our models provide a good match to these age trends. Figure \ref{fig:yield-outflow} illustrates a point that will arise as a theme throughout the remainder of this paper, which is that the signatures of a substantial dilution event and subsequent re-enrichment, characteristic of the two-infall scenario, are simply not present in the observed age--abundance trends.

\todo{Contrast with: multiple sequences in Anders et al. (2023), Xiang \& Rix (2022), Sahlholdt et al. (2022).}

@@

>>>>

\begin{figure}
    \centering
    \includegraphics[width=\onecolumn]{\figpath{gas_abundance_evolution.pdf}}
    \caption{The ISM abundance evolution at $R_{\rm gal}=8\kpc$ of three multi-zone models at different yield scales (see Table \ref{tab:yields}), each of which feature a major dilution event 9 Gyr ago.
    % The dark gray (light gray) shaded region plots the 68\textsuperscript{th} (95\textsuperscript{th}) percentile of APOGEE abundances in the Solar neighborhood ($7\leq R_{\rm gal}<9\kpc$, $|z|<0.5\kpc$) with NN ages from \citet{leung_variational_2023}. The black line (black points) plots the median (mode) of the abundance data in 1 Gyr-wide age bins, and the gray error bars along the bottom of each panel indicate the median age and abundance errors as a function of age. 
    \added{The 2-D histograms show the APOGEE stellar age--abundance distributions in the Solar neighborhood ($7\leq R_{\rm gal}<9\kpc$, $|z|<0.5\kpc$), adopting the \citet{leung_variational_2023} ages. The black curve (points) plots the rolling median (binned mode) of the abundance data as a function of age, and the gray error bars along the bottom of each panel indicate the median age and abundance errors as a function of age.}
    The left-hand marginal panels show the predicted (blue, pink, and green) and observed (gray) stellar abundance distributions \added{in the same region}, which are boxcar-smoothed with a width of 0.05 dex for visual clarity.}
    \label{fig:yield-outflow}
\end{figure}

Figure \ref{fig:yield-outflow} shows the \added{APOGEE age--abundance distributions at the Solar annulus, as well as the binned mode abundance and rolling median as a function of age}, using the NN ages from \citet{leung_variational_2023}. As shown by \citet{johnson_milky_2025}, the peak of the MDF is less susceptible to modification by radial migration, making it a more reliable proxy for ISM chemistry at a given lookback time than the mean or median. \added{While the age--metallicity distribution is clumpy in places (see Section \ref{sec:amr}), we observe that ${\rm mode}(\mathOH)\gtrsim0$ for the past 10 Gyr.}

We over-plot the ISM abundance evolution in our $y/Z_\odot=1,2,$ and $3$ models with the fiducial parameters at $R_{\rm gal}=8\kpc$ for comparison. The models are in broad agreement with each other and with the data at lookback times of $\lesssim5\Gyr$. At ages of $\sim5-8\Gyr$, the models underpredict the metallicities of the observed stars by up to half an order of magnitude, a natural consequence of the dilution associated with the second infall event. This discrepancy is modestly reduced in the higher $y/Z_\odot$ models, which re-enrich more rapidly. However, these models exacerbate tensions with the age--[O/Fe] relation because the O abundance increases more before enrichment from second-infall SNe Ia becomes important, making the $\alpha$-enhancement from the ensuing starburst more pronounced. Overall, none of our models provide a good match to these age trends. Figure \ref{fig:yield-outflow} illustrates a point that will arise as a theme throughout the remainder of this paper, which is that the signatures of a substantial dilution event and subsequent re-enrichment, characteristic of the two-infall scenario, are simply not present in the observed age--abundance trends.

@@

\subsection{The Age--Metallicity Relation}
\label{sec:amr}

\added{A few recent studies have reported the discovery of multiple disconnected sequences in the local age--metallicity relation \citep[e.g.,][]{sahlholdt_characterizing_2022,xiang_time-resolved_2022,anders_spectroscopic_2023}. \citet{nissen_high-precision_2020} argued that these sequences provide evidence for the two-infall scenario, although \citet{plotnikova_chemical_2024} showed that multiple sequences can arise due to selection effects, and \citet{chen_recent_2025} showed that radial migration can also lead to such patterns. There are hints of multiple sequences in Figure \ref{fig:yield-outflow}, where a break in the distribution at ages of $\sim5\Gyr$ coincides with a $\sim0.2\dex$ drop in [O/H] and [Fe/H]. The sequences are less pronounced in the \citet{leung_variational_2023} catalog than reported by \citet{anders_spectroscopic_2023}, likely related to a systematic offset between the two catalogs around the same age \citep[see Figure 6 of][]{anders_spectroscopic_2023}. On the other hand, the [C/N] age-metallicity relation of \citet{roberts_cn_2025} is even flatter and does not appear to have multiple sequences. Even if these sequences are real, we find it difficult to explain them using the two-infall scenario\textemdash the observed dilution is much smaller than predicted by any of our models, and the timing of the split between the sequences is too late for the birth of the thin disk, although it is interestingly close to the first pericentric passage of the Sagittarius dwarf galaxy \citep{ruiz-lara_recurrent_2020}.}
\todo{Mention APOGEE selection function?}
% Anders+ use APOGEE clump stars. Their split is 5-6 Gyr ago
% Xiang & Rix use Gaia subgiants. Their split is 8 Gyr ago
% Sahlholdt+ use GALAH turnoff stars. Their split is 7-9 Gyr ago.

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{\figpath{compare_age_catalogs.pdf}}
    \caption{Comparison between NN- and [C/N]-derived ages for our sample.}
    \label{fig:compare-age-catalogs}
\end{figure*}
